---
title: "Rhode Island 2026 SGP Analyses"
subtitle: "RICAS & SAT Student Growth Percentile Results"
author:
  - name: "Damian W. Betebenner"
    affiliation: "National Center for the Improvement of Educational Assessment"
date: 2026-08-27
date-format: "[RIDE TAC Meeting · ]MMMM D, YYYY"
format:
  revealjs:
    output-file: index.html
    theme: [default, styles/presentation-theme.scss]
    slide-number: true
    transition: fade
    transition-speed: fast
    width: 1920
    height: 1080
    margin: 0.06
    center: false
    hash: true
    history: true
    controls: true
    progress: true
    html-math-method: mathml
    fig-align: center
    embed-resources: true
    include-in-header:
      - text: |
          <link rel="preload" href="assets/fonts/NotoSansMath.woff2" as="font" type="font/woff2" crossorigin>
          <style>
            @font-face {
              font-family: "Noto Sans Math";
              src: url("assets/fonts/NotoSansMath.woff2") format("woff2");
              font-display: swap;
            }
          </style>
    include-after-body:
      - text: |
          <script src="styles/presentation-figure-modal.js"></script>
execute:
  echo: false
  warning: false
  message: false
knitr:
  opts_chunk:
    dev: "svg"
    fig.align: "center"
---

```{r setup}
#| include: false

source("R/00_helpers.R")
suppressPackageStartupMessages({
  library(data.table)
  library(ggplot2)
})

## ---- Load the compact summary cache (never the 70MB LONG file) -------------
S   <- readRDS("summary/ri_sgp_summary.rds")
ov  <- as.data.table(S$overall)
lev <- as.data.table(S$levels)
sg  <- as.data.table(S$subgroups)
sgp_pool <- as.data.table(S$subgroups_pooled)
tg  <- as.data.table(S$targets)
tgh <- as.data.table(S$targets_highneed)
M   <- S$meta

AY  <- M$analysis_year          # "2025_2026"
AYL <- M$analysis_year_label    # "2026"

## ---- Palette + ggplot theme -----------------------------------------------
COHORT_COL   <- "#3498DB"
BASELINE_COL <- "#E67E22"
LEVEL_COLS   <- c(Low = "#C0392B", Typical = "#95A5A6", High = "#27AE60")
GRID_COL     <- "#D8D8D4"
TEXT_COL     <- "#141410"

theme_ri <- function(base_size = 22) {
  theme_minimal(base_size = base_size, base_family = "Noto Sans") +
    theme(
      text            = element_text(color = TEXT_COL),
      plot.title      = element_text(face = "bold", size = rel(1.05)),
      plot.subtitle   = element_text(color = "#5c5c58"),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = GRID_COL, linewidth = 0.4),
      strip.text      = element_text(face = "bold"),
      legend.position = "top",
      legend.title    = element_blank(),
      axis.title      = element_text(color = "#5c5c58")
    )
}

## ---- Lookups / formatting helpers used inline throughout -------------------
mmn_cell <- function(content, grade_label, norm = c("cohort", "baseline"), year = AY) {
  norm <- match.arg(norm)
  r <- ov[CONTENT_AREA == content & GRADE_LABEL == grade_label & YEAR == year]
  if (!nrow(r) || (norm == "cohort" && r$n_cohort == 0)) return("&mdash;")
  if (norm == "cohort") fmt_mmn(r$mean_cohort, r$median_cohort, r$n_cohort)
  else fmt_mmn(r$mean_baseline, r$median_baseline, r$n_baseline)
}

get_val <- function(content, grade_label, col, year = AY) {
  r <- ov[CONTENT_AREA == content & GRADE_LABEL == grade_label & YEAR == year]
  if (!nrow(r)) return(NA_real_)
  r[[col]]
}

## ---- Goodness-of-fit figure helpers (clean vendored SVGs) ------------------
gof_token <- function(content_area) {
  data.table::fcase(
    content_area == "ELA", "ELA",
    content_area == "MATHEMATICS", "MATH",
    content_area == "ELA_SAT", "ELA_SAT",
    content_area == "MATHEMATICS_SAT", "MATH_SAT",
    default = content_area
  )
}

gof_src <- function(content_area, grade, year, norm = "cohort") {
  file.path("Figures", "gof",
            sprintf("%s_%s_%s_g%s.svg", gof_token(content_area), year, norm, grade))
}

## One <figure> (works with the theme figure-modal lightbox: `.reveal figure img`)
gof_figure <- function(content_area, grade, year, norm = "cohort",
                       caption = NULL, alt = NULL) {
  src <- gof_src(content_area, grade, year, norm)
  if (!file.exists(src)) {
    return(sprintf('<div style="padding:1em;color:#c0392b;font-size:0.6em">missing: %s</div>', src))
  }
  if (is.null(caption)) caption <- sprintf("%s Grade %s &mdash; %s (%s)",
                                           pretty_subject(content_area), grade,
                                           pretty_year(year), norm)
  if (is.null(alt)) alt <- gsub("&mdash;", "-", caption)
  sprintf(
    '<figure style="margin:0;text-align:center">
       <img src="%s" alt="%s" style="width:100%%;height:auto"/>
       <figcaption>%s</figcaption>
     </figure>', src, alt, caption)
}

## Emit raw HTML as a pandoc passthrough block so `native_divs` does not try to
## pair our <div>s with Quarto's auto `::: {.cell}` wrapper (which otherwise
## leaks stray "::" text). Use for every asis chunk that prints raw HTML.
emit_html <- function(x) cat("\n```{=html}\n", x, "\n```\n", sep = "")

## Responsive grid of GoF figures
gof_grid <- function(figs, min_col = "300px", gap = "0.6em") {
  emit_html(sprintf(
    '<div style="display:grid;grid-template-columns:repeat(auto-fit,minmax(%s,1fr));gap:%s;align-items:start">%s</div>',
    min_col, gap, paste(figs, collapse = "")))
}
```

## The next 20 minutes {.center}

::: {.visually-hidden}
Roadmap for the talk
:::

::: {style="font-size: 1.05em;"}
1. **What was run** &mdash; RICAS and PSAT/SAT analyses, cohort and baseline
2. **2026 results** &mdash; overall growth and model fit
3. **Trends over time** &mdash; pandemic recovery and possible scale drift
4. **Growth equity** &mdash; results by student group
5. **Target attainment** &mdash; catch-up / keep-up and move-up / stay-up
6. **Questions for the TAC**
:::

::: {.key-insight}
**`r fmt_count(M$n_growth_cohort_total)`** students received a `r AYL` growth percentile across RICAS (grades 4&ndash;8) and SAT (grade 11), in both cohort-referenced and pre-pandemic baseline-referenced norms.
:::

::: {.notes}
This is a TAC audience &mdash; no SGP primer. Set the arc, then move. All numbers on the slides are generated directly from the LONG data via a cached summary, so they are current as of the latest data run.
:::


# SGP results for 2026 {.center}

::: {.visually-hidden}
Section one: 2026 results
:::


## What was run: RICAS (grades 4&ndash;8)

Two SGP analyses were conducted on the `r AYL` RICAS data (ELA and Mathematics):

- **Cohort-referenced** &mdash; growth norms built from the most recent cohorts; by construction these center mean/median SGP on **50**.
- **Baseline-referenced** &mdash; historical (pre-pandemic 2017&ndash;2019) coefficient matrices applied to current data. These are **not** centered on 50 and let us examine COVID-era academic recovery and possible **scale drift**.
- **Projections** (lagged and straight) were calculated using both norm sets.

::: {.key-insight}
Cohort norms answer *"how did this student grow relative to academic peers this year?"* Baseline norms answer *"how does that growth compare to pre-pandemic rates of learning?"*
:::

::: {.notes}
Same two-analysis structure as prior years. Emphasize that baseline is the diagnostic lens for recovery / drift.
:::


## What was run: PSAT &rarr; SAT (grade 11)

Two SGP analyses were conducted on the PSAT-10 &rarr; SAT progression (single prior):

- **Cohort-referenced** &mdash; centered on 50 by construction.
- **Baseline-referenced** &mdash; pre-pandemic (2018&ndash;2019) norms applied to current data.
- **No projections** &mdash; SAT (grade 11) is the last tested grade in Rhode Island.

::: {.theorem-box}
::: {.theorem-box-title}
Single-prior growth
:::
Grade-11 growth uses one prior (PSAT-10). It is a shorter growth window than the RICAS grades, so baseline comparisons are correspondingly more sensitive to form and linking effects.
:::

::: {.notes}
Flag the single-prior caveat now; it matters when we read the SAT baseline numbers.
:::


## Provenance &amp; reproducibility {.smaller}

::: {.columns}
::: {.column width="56%"}
- All analyses used data supplied by the Rhode Island Department of Education.
- Conducted with the **`SGP` R package** (version `r M$sgp_package_version`).
- All analyses converged; goodness-of-fit plots confirm successful analyses.
- Source code is public on GitHub:
  - Analyses: [github.com/CenterForAssessment/Rhode_Island](https://github.com/CenterForAssessment/Rhode_Island)
  - SGP package: [github.com/CenterForAssessment/SGP](https://github.com/CenterForAssessment/SGP)
:::
::: {.column width="42%"}
::: {.key-insight-box}
::: {.key-insight-box-title}
Reproducible pipeline
:::
This deck is generated from a cached summary of the student-level LONG data. When RIDE supplies revised data, re-running the pipeline refreshes **every** number and table automatically.
:::
:::
:::

::: {.notes}
The reproducibility box is new this year &mdash; it previews the "I'll re-run this when the revised file arrives" workflow.
:::


## 2026 SGP results at a glance

```{r results-table}
#| output: asis

grades <- c("4", "5", "6", "7", "8", "11 (SAT)")
content_ela  <- function(g) if (g == "11 (SAT)") "ELA_SAT" else "ELA"
content_math <- function(g) if (g == "11 (SAT)") "MATHEMATICS_SAT" else "MATHEMATICS"

rows <- vapply(grades, function(g) {
  sprintf(
    '<tr><th class="rt-grade">%s</th><td>%s</td><td class="rt-base">%s</td><td>%s</td><td class="rt-base">%s</td></tr>',
    g,
    mmn_cell(content_ela(g),  g, "cohort"),
    mmn_cell(content_ela(g),  g, "baseline"),
    mmn_cell(content_math(g), g, "cohort"),
    mmn_cell(content_math(g), g, "baseline")
  )
}, character(1))

emit_html(sprintf('
<style>
  table.results-table { border-collapse: collapse; width: 100%%; font-size: 0.72em; }
  table.results-table th, table.results-table td { padding: 0.35em 0.6em; text-align: center; border-bottom: 1px solid var(--color-border); }
  table.results-table thead .grp { font-family: "Josefin Sans", sans-serif; font-weight: 600; }
  table.results-table thead .grp-ela  { color: %s; border-bottom: 2px solid %s; }
  table.results-table thead .grp-math { color: %s; border-bottom: 2px solid %s; }
  table.results-table .rt-grade { text-align: left; font-family: "Josefin Sans", sans-serif; }
  table.results-table .rt-base { color: %s; }
  table.results-table tbody tr:hover { background: var(--color-surface-subtle); }
  .rt-caption { font-size: 0.55em; color: var(--color-secondary); font-style: italic; margin-top: 0.4em; }
</style>
<table class="results-table">
  <thead>
    <tr>
      <th rowspan="2">Grade</th>
      <th class="grp grp-ela" colspan="2">ELA</th>
      <th class="grp grp-math" colspan="2">Mathematics</th>
    </tr>
    <tr>
      <th>Cohort</th><th class="rt-base">Baseline</th>
      <th>Cohort</th><th class="rt-base">Baseline</th>
    </tr>
  </thead>
  <tbody>%s</tbody>
</table>
<div class="rt-caption">Each cell: mean&#8202;/&#8202;median&#8202;/&#8202;N of student growth percentiles, %s. Cohort norms center on 50 by construction; baseline norms are referenced to pre-pandemic rates of learning.</div>
',
COHORT_COL, COHORT_COL, "#7a5c33", "#7a5c33", BASELINE_COL,
paste(rows, collapse = "\n"), AYL))
```

::: {.notes}
Walk the cohort columns first (all ~50, as expected), then the baseline columns &mdash; that is where the story is. The deviations from 50 are what we dig into next.
:::


## Reading the 2026 baseline results

```{r baseline-narrative}
#| output: asis

## Flag baseline means that deviate meaningfully from 50 (analysis year, RICAS + SAT).
by <- ov[YEAR == AY & n_baseline > 0]
by[, dev := mean_baseline - 50]
flag <- by[abs(dev) >= 2][order(-abs(dev))]

mk <- function(x) sprintf("%s Grade %s (mean <strong>%.1f</strong>)",
                          x$SUBJECT, x$GRADE_LABEL, x$mean_baseline)
li <- function(dt) {
  if (!nrow(dt)) return("<li>None materially different</li>")
  paste0("<li>", vapply(seq_len(nrow(dt)), function(i) mk(dt[i]), character(1)), "</li>", collapse = "")
}
above <- flag[dev > 0]; below <- flag[dev < 0]

emit_html(sprintf('
<div style="display:flex;gap:2em;align-items:flex-start">
  <div style="flex:1">
    <p style="font-weight:600;color:#27ae60;margin-bottom:0.2em">Above pre-pandemic rates (baseline mean &gt; 52)</p>
    <ul>%s</ul>
  </div>
  <div style="flex:1">
    <p style="font-weight:600;color:#c0392b;margin-bottom:0.2em">Below pre-pandemic rates (baseline mean &lt; 48)</p>
    <ul>%s</ul>
  </div>
</div>', li(above), li(below)))
```

::: {.key-insight}
Cohort-referenced results are, as expected, centered on 50 with excellent model fit. The **baseline** results are uneven across grade/content &mdash; the pattern we investigate in the trends section.
:::

::: {.notes}
This slide is generated from the data, so the flagged grades update automatically with a new run. Keep interpretation cautious: deviations can reflect real growth change, cohort composition, or scale drift.
:::


## Growth distribution: Low / Typical / High

```{r level-distribution}
#| fig-width: 16
#| fig-height: 6.2

ld <- lev[YEAR == AY & !is.na(LEVEL)]
ld[, GRADE_LABEL := factor(GRADE_LABEL, levels = c("4","5","6","7","8","11 (SAT)"))]
ld[, NORM := factor(NORM, levels = c("Cohort", "Baseline"))]

ggplot(ld, aes(x = GRADE_LABEL, y = pct, fill = LEVEL)) +
  geom_col(width = 0.78) +
  geom_hline(yintercept = c(33.3, 66.7), color = "white", linewidth = 0.5, linetype = "22") +
  facet_grid(SUBJECT ~ NORM) +
  scale_fill_manual(values = LEVEL_COLS) +
  scale_y_continuous(labels = function(x) paste0(x, "%"), expand = expansion(mult = c(0, 0.02))) +
  labs(x = NULL, y = NULL,
       title = paste0("Share of students at each growth level, ", AYL),
       subtitle = "Cohort norms split roughly one-third each; baseline shifts reveal recovery / drift") +
  theme_ri(20)
```

::: {.notes}
Cohort panels should show the classic ~1/3 Low, Typical, High. Where baseline panels tilt toward Low or High is where pre-pandemic-referenced growth diverges. Point to the largest tilts.
:::


# Goodness of fit {.center}

::: {.visually-hidden}
Section two: goodness of fit
:::

::: {.notes}
For a TAC audience this is a fit-confirmation section. Keep it brisk; the plots are click-to-enlarge if anyone wants detail.
:::


## RICAS ELA &mdash; cohort model fit, `r AYL`

```{r gof-ela-cohort}
#| output: asis
gof_grid(vapply(c("4","5","6","7","8"),
                function(g) gof_figure("ELA", g, AY, "cohort",
                                       caption = sprintf("ELA Grade %s", g)),
                character(1)))
```

::: {.notes}
Uniform bands across prior achievement = good fit. All five grades converge cleanly.
:::


## RICAS Mathematics &mdash; cohort model fit, `r AYL`

```{r gof-math-cohort}
#| output: asis
gof_grid(vapply(c("4","5","6","7","8"),
                function(g) gof_figure("MATHEMATICS", g, AY, "cohort",
                                       caption = sprintf("Mathematics Grade %s", g)),
                character(1)))
```


## SAT &mdash; cohort model fit, `r AYL`

```{r gof-sat-cohort}
#| output: asis
gof_grid(c(
  gof_figure("ELA_SAT", "EOCT", AY, "cohort", caption = "ELA &mdash; SAT (grade 11)"),
  gof_figure("MATHEMATICS_SAT", "EOCT", AY, "cohort", caption = "Mathematics &mdash; SAT (grade 11)")
), min_col = "420px")
```

::: {.notes}
Single-prior SAT fit. Note the shorter growth window relative to RICAS.
:::


## Cohort vs. baseline fit &mdash; ELA Grade 7

```{r gof-contrast}
#| output: asis
gof_grid(c(
  gof_figure("ELA", "7", AY, "cohort",   caption = "ELA Grade 7 &mdash; Cohort norms"),
  gof_figure("ELA", "7", AY, "baseline", caption = "ELA Grade 7 &mdash; Baseline norms")
), min_col = "420px")
```

::: {.key-insight}
Both norm sets fit the data well. The difference is the **reference**: cohort centers on this year's peers; baseline holds the pre-pandemic standard fixed &mdash; so a level shift here is signal, not misfit.
:::


# In-depth trend analysis {.center}

::: {.visually-hidden}
Section three: trends over time
:::

::: {.notes}
The analytic heart of the talk: pandemic recovery vs. scale drift. Slow down here.
:::


## Pandemic-related academic recovery

::: {.columns}
::: {.column width="54%"}
Five years on from the onset of the pandemic, we examine **changes in growth over time** to distinguish among:

- **Ripple effects** &mdash; continued downstream impact for students whose early schooling was disrupted.
- **Scale drift / linking** &mdash; movement in the underlying scale over time.
- **Real change** in student growth.
:::
::: {.column width="42%"}
::: {.theorem-box}
::: {.theorem-box-title}
Two views of the same data
:::
**Diagonal** &mdash; follow one cohort as it ages grade-to-grade.

**Vertical** &mdash; hold a grade fixed and watch it move year-to-year (a drift indicator).
:::
:::
:::

::: {.notes}
Set up the two reading directions we'll use on the next slides.
:::


## Where were they during the pandemic?

```{r pandemic-grid}
#| output: asis
## Grade in 2026 -> grade during the two pandemic school years.
tbl <- data.frame(
  cur   = c("Grade 4","Grade 5","Grade 6","Grade 7","Grade 8","Grade 11 (SAT)"),
  s2020 = c("PK","K","Grade 1","Grade 2","Grade 3","Grade 6"),
  s2021 = c("K","Grade 1","Grade 2","Grade 3","Grade 4","Grade 7")
)
rows <- apply(tbl, 1, function(r)
  sprintf('<tr><th style="text-align:left;font-family:\'Josefin Sans\',sans-serif">%s</th><td>%s</td><td>%s</td></tr>',
          r["cur"], r["s2020"], r["s2021"]))
emit_html(sprintf('
<table class="results-table" style="font-size:0.7em">
  <thead><tr><th>Current (2026)</th><th>Spring 2020</th><th>2020&ndash;2021</th></tr></thead>
  <tbody>%s</tbody>
</table>
<div class="rt-caption">The lower grades tested in 2026 spent their earliest, most foundational schooling years during pandemic disruption.</div>',
paste(rows, collapse = "")))
```

::: {.notes}
The grade-7 and grade-8 cohorts (early-literacy grades during the pandemic) are the ones we track next.
:::


## Trend: same grade over time (drift indicator)

```{r trend-samegrade}
#| fig-width: 16
#| fig-height: 6.2

tr <- ov[ASSESSMENT == "RICAS" & n_baseline > 0 &
         YEAR %in% c("2022_2023","2023_2024","2024_2025","2025_2026")]
tr[, GRADE := factor(GRADE, levels = as.character(4:8))]

ggplot(tr, aes(x = YEAR_LABEL, y = mean_baseline, color = GRADE, group = GRADE)) +
  geom_hline(yintercept = 50, color = "#9a9a94", linetype = "22") +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.6) +
  facet_wrap(~ SUBJECT) +
  scale_color_brewer(palette = "Dark2", name = "Grade") +
  labs(x = NULL, y = "Mean baseline SGP",
       title = "Baseline-referenced growth by grade, over time",
       subtitle = "Large year-to-year moves for a fixed grade are consistent with scale drift; 50 = pre-pandemic rate") +
  theme_ri(20) + theme(legend.position = "right")
```

::: {.notes}
This is the vertical/drift reading, fully data-driven. Point to grades whose baseline mean swings across years &mdash; those are drift candidates.
:::


## ELA Grade 7 over time &mdash; goodness of fit

```{r gof-drift-ela7}
#| output: asis
gof_grid(vapply(c("2022_2023","2023_2024","2024_2025","2025_2026"),
                function(y) gof_figure("ELA", "7", y, "baseline",
                                       caption = sprintf("ELA Grade 7 &mdash; %s", pretty_year(y))),
                character(1)), min_col = "300px")
```

::: {.notes}
Same grade, four administrations, baseline norms. Consistent shape = stable; systematic drift in the bands = a linking/scale concern.
:::


## Mathematics Grade 6 over time &mdash; goodness of fit

```{r gof-drift-math6}
#| output: asis
gof_grid(vapply(c("2022_2023","2023_2024","2024_2025","2025_2026"),
                function(y) gof_figure("MATHEMATICS", "6", y, "baseline",
                                       caption = sprintf("Math Grade 6 &mdash; %s", pretty_year(y))),
                character(1)), min_col = "300px")
```


## Diagonal: following the 2026 Grade-8 cohort

```{r gof-diagonal-ela8}
#| output: asis
gof_grid(c(
  gof_figure("ELA", "6", "2023_2024", "baseline", caption = "Grade 6 &mdash; 2024"),
  gof_figure("ELA", "7", "2024_2025", "baseline", caption = "Grade 7 &mdash; 2025"),
  gof_figure("ELA", "8", "2025_2026", "baseline", caption = "Grade 8 &mdash; 2026")
), min_col = "340px")
```

::: {.key-insight}
The same students, three consecutive years (ELA). This cohort was in the early-literacy grades during the pandemic &mdash; the diagonal lets us watch their baseline-referenced growth as they progressed.
:::

::: {.notes}
Diagonal reading: this is one cohort aging through the system, not a fixed grade. Contrast with the previous drift slides.
:::


# Growth equity {.center}

::: {.visually-hidden}
Section four: growth by student group
:::

::: {.notes}
Descriptive equity lens: are growth rates comparable across groups? Keep it factual.
:::


## Growth by student group (RICAS, `r AYL`)

```{r equity-table}
#| output: asis

dims <- c("English Learner", "Students w/ Disability", "Economically Disadv.")
sp <- sgp_pool[ASSESSMENT == "RICAS" & DIMENSION %in% dims]

cell <- function(subj, grp, col) {
  r <- sp[SUBJECT == subj & GROUP == grp]
  if (!nrow(r)) return("&mdash;")
  fmt_mmn(r[[paste0("mean_", col)]], r[[paste0("median_", col)]], r[[paste0("n_", col)]])
}

body <- ""
for (d in dims) {
  grps <- sp[DIMENSION == d, unique(GROUP)]
  body <- paste0(body, sprintf('<tr class="dim-row"><th colspan="5">%s</th></tr>', d))
  for (grp in grps) {
    body <- paste0(body, sprintf(
      '<tr><th class="rt-grade">%s</th><td>%s</td><td class="rt-base">%s</td><td>%s</td><td class="rt-base">%s</td></tr>',
      grp, cell("ELA", grp, "cohort"), cell("ELA", grp, "baseline"),
      cell("Mathematics", grp, "cohort"), cell("Mathematics", grp, "baseline")))
  }
}

emit_html(sprintf('
<style>
  table.results-table tr.dim-row th { text-align:left; background: var(--color-surface-subtle);
    font-family:"Josefin Sans",sans-serif; color: var(--color-primary); padding-top:0.5em; }
</style>
<table class="results-table" style="font-size:0.62em">
  <thead>
    <tr><th rowspan="2">Group</th>
        <th class="grp grp-ela" colspan="2">ELA</th>
        <th class="grp grp-math" colspan="2">Mathematics</th></tr>
    <tr><th>Cohort</th><th class="rt-base">Baseline</th><th>Cohort</th><th class="rt-base">Baseline</th></tr>
  </thead>
  <tbody>%s</tbody>
</table>
<div class="rt-caption">mean&#8202;/&#8202;median&#8202;/&#8202;N, RICAS grades 4&ndash;8 pooled, %s.</div>', body, AYL))
```

::: {.notes}
Cohort SGPs near 50 across groups indicate comparable within-year growth. Baseline gaps reflect pre-pandemic-referenced differences. Note where cohort and baseline tell different stories.
:::


## Equity gaps at a glance

```{r equity-plot}
#| fig-width: 16
#| fig-height: 6.0

dims <- c("English Learner", "Students w/ Disability", "Economically Disadv.")
ep <- sgp_pool[ASSESSMENT == "RICAS" & DIMENSION %in% dims,
               .(mean_sgp = weighted.mean(mean_cohort, n_cohort),
                 mean_base = weighted.mean(mean_baseline, n_baseline)),
               by = .(DIMENSION, GROUP)]
epl <- melt(ep, id.vars = c("DIMENSION","GROUP"),
            measure.vars = c("mean_sgp","mean_base"),
            variable.name = "norm", value.name = "sgp")
epl[, norm := factor(norm, labels = c("Cohort", "Baseline"))]
epl[, GROUP := gsub(" \\(.*\\)$", "", GROUP)]

ggplot(epl, aes(x = sgp, y = reorder(GROUP, sgp), fill = norm)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.65) +
  geom_vline(xintercept = 50, color = "#9a9a94", linetype = "22") +
  facet_wrap(~ DIMENSION, scales = "free_y", ncol = 1) +
  scale_fill_manual(values = c(Cohort = COHORT_COL, Baseline = BASELINE_COL)) +
  labs(x = "Mean SGP (RICAS ELA + Math pooled)", y = NULL,
       title = paste0("Growth by student group, ", AYL)) +
  theme_ri(19)
```

::: {.notes}
The dashed line is 50. Cohort bars near 50 = comparable current-year growth; the cohort-to-baseline distance shows the pre-pandemic-referenced gap. Descriptive, not causal.
:::


# Target attainment {.center}

::: {.visually-hidden}
Section five: catch-up / keep-up
:::


## Catch-up / keep-up attainment (RICAS, `r AYL`)

```{r targets-plot}
#| fig-width: 16
#| fig-height: 6.0

ct <- tg[METRIC == "Catch Up / Keep Up" & NORM == "Cohort" & ASSESSMENT == "RICAS"]
ct[, GRADE_LABEL := factor(GRADE_LABEL, levels = as.character(4:8))]
ct[, aspiration := factor(aspiration, levels = c("Catch Up", "Keep Up"))]

ggplot(ct, aes(x = GRADE_LABEL, y = rate, fill = aspiration)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.68) +
  facet_wrap(~ SUBJECT) +
  scale_fill_manual(values = c("Catch Up" = "#8E44AD", "Keep Up" = COHORT_COL)) +
  scale_y_continuous(labels = function(x) paste0(x, "%"),
                     limits = c(0, 100), expand = expansion(mult = c(0, 0.02))) +
  labs(x = "Grade", y = "% attaining target",
       title = "Share of students on track to catch up / keep up",
       subtitle = "Catch Up = students below proficient; Keep Up = students at/above proficient (cohort targets)") +
  theme_ri(20)
```

::: {.notes}
"Catch Up: Yes" among students who need to catch up; "Keep Up: Yes" among those at/above. Keep-up rates run higher than catch-up rates &mdash; the recovery challenge is concentrated in catch-up.
:::


## Target attainment by prior achievement

```{r targets-highneed}
#| output: asis

th <- tgh[NORM == "Cohort"]
mk_stat <- function(status_label, aspiration_label) {
  r <- th[grepl(sub(" .*", "", status_label), HIGH_NEED_STATUS) & aspiration == aspiration_label]
  if (!nrow(r)) return("&mdash;")
  sprintf("%.0f%%", r$rate)
}

low  <- th[grepl("< 25", HIGH_NEED_STATUS) & aspiration == "Catch Up"]
high <- th[grepl("> 75", HIGH_NEED_STATUS) & aspiration == "Keep Up"]

emit_html(sprintf('
<div style="display:flex;gap:2em;justify-content:center;margin-top:0.5em">
  <div class="stat-callout"><div class="stat-number">%.0f%%</div>
    <div class="stat-label">of lowest-prior-achievement students (&lt;25th %%ile)<br/>are on track to <strong>catch up</strong></div></div>
  <div class="stat-callout"><div class="stat-number">%.0f%%</div>
    <div class="stat-label">of highest-prior-achievement students (&gt;75th %%ile)<br/>are on track to <strong>keep up</strong></div></div>
</div>', low$rate, high$rate))
```

::: {.key-insight}
Students entering below the 25th percentile face the steepest catch-up targets and attain them least often &mdash; the central equity concern for accountability use of these growth results.
:::

::: {.notes}
Numbers pull from the high-need contrast in the cache. This frames the accountability-use conversation.
:::


## Summary

- **Cohort-referenced SGPs** show excellent model fit across RICAS and SAT &mdash; suitable for accountability calculations.
- **Baseline-referenced SGPs** are uneven: some grade/content pairs sit above pre-pandemic rates of learning, others below.
- **Trend analysis** (same-grade-over-time and diagonal-cohort) shows changes that are difficult to attribute to a single mechanism &mdash; recovery, cohort composition, and scale drift all remain in play.
- **Continued equating / linking effects over time cannot be ruled out** as a contributor to the baseline movement.
- **Equity and target-attainment** results locate the recovery challenge primarily in catch-up for the lowest-prior-achievement students.

::: {.notes}
Land the honest bottom line: cohort norms are solid; baseline movement is real but multiply-determined, and drift is not ruled out.
:::


## Questions for the TAC {.smaller}

::: {.key-insight-box}
::: {.key-insight-box-title}
We would value the committee's guidance on
:::
1. **Disentangling mechanisms.** What additional analyses would help separate pandemic ripple effects, scale drift, and genuine changes in student growth over time?
2. **Baseline norm maintenance.** If baseline SGPs continue going forward, should Rhode Island **re-establish baseline norms** (moving off the pre-pandemic 2017&ndash;2019 reference) &mdash; and on what cadence?
3. **Form effects.** Would examining cohort-referenced growth **by form** (potentially leveraging MA data) be a productive way to probe linking/form effects? Is that something MA would be interested in pursuing?
4. **Accountability use.** Given the baseline unevenness, are there guardrails the TAC would recommend for interpreting baseline-referenced growth in public reporting?
:::

::: {.notes}
This is the discussion slide &mdash; leave it up and open the floor. These carry forward last year's open questions and add the reporting-guardrails item.
:::


## Thank you {.center}

::: {style="text-align: center; padding-top: 1em;"}
::: {.conference-badge}
RIDE Technical Advisory Committee
:::

**Rhode Island 2026 SGP Analyses**

::: {style="font-size: 0.7em; color: #7c7c7c;"}
Damian W. Betebenner &middot; National Center for the Improvement of Educational Assessment
:::
:::

::: {.notes}
Close and thank the committee. Offer to circulate the deck and the reproducible pipeline.
:::
