---
title: "Rhode Island 2026 SGP Analyses"
subtitle: "RICAS & SAT Student Growth Percentile Results<br/><span class='revision-badge'>Revision 2 &middot; revised following TAC feedback</span>"
author:
  - name: "Damian W. Betebenner"
    affiliation: "National Center for the Improvement of Educational Assessment"
date: 2026-08-27
date-format: "[RIDE TAC Meeting · ]MMMM D, YYYY"
format:
  revealjs:
    output-file: index.html
    theme: [default, styles/presentation-theme.scss]
    slide-number: true
    transition: fade
    transition-speed: fast
    width: 1920
    height: 1080
    margin: 0.06
    center: false
    hash: true
    history: true
    controls: true
    progress: true
    html-math-method: mathml
    fig-align: center
    embed-resources: true
    include-in-header:
      - text: |
          <link rel="preload" href="assets/fonts/NotoSansMath.woff2" as="font" type="font/woff2" crossorigin>
          <style>
            @font-face {
              font-family: "Noto Sans Math";
              src: url("assets/fonts/NotoSansMath.woff2") format("woff2");
              font-display: swap;
            }
            .revision-badge {
              display: inline-block;
              margin-top: 0.6em;
              padding: 0.18em 0.7em;
              border: 1px solid currentColor;
              border-radius: 2em;
              font-size: 0.5em;
              letter-spacing: 0.04em;
              text-transform: uppercase;
              opacity: 0.85;
            }
            .tac-tag {
              display: inline-block;
              font-size: 0.62em;
              letter-spacing: 0.03em;
              text-transform: uppercase;
              color: #7a5c33;
              border-left: 3px solid #E67E22;
              padding-left: 0.5em;
              margin-bottom: 0.35em;
            }
          </style>
    include-after-body:
      - text: |
          <script src="styles/presentation-figure-modal.js"></script>
execute:
  echo: false
  warning: false
  message: false
knitr:
  opts_chunk:
    dev: "svg"
    fig.align: "center"
---

```{r setup}
#| include: false

source("R/00_helpers.R")
suppressPackageStartupMessages({
  library(data.table)
  library(ggplot2)
})

## ---- Load the compact summary cache (never the 70MB LONG file) -------------
S   <- readRDS("summary/ri_sgp_summary.rds")
ov  <- as.data.table(S$overall)
lev <- as.data.table(S$levels)
sg  <- as.data.table(S$subgroups)
sgp_pool <- as.data.table(S$subgroups_pooled)
tg  <- as.data.table(S$targets)
tgh <- as.data.table(S$targets_highneed)
## Added in the TAC revision
xsec <- as.data.table(S$xsec)
ppc  <- as.data.table(S$pp)
prof <- as.data.table(S$proficiency)
sgpd <- as.data.table(S$sgp_distribution)
M   <- S$meta
BNY <- if (!is.null(M$baseline_norm_year)) M$baseline_norm_year else "2018_2019"
BNYL <- pretty_year(BNY)

AY  <- M$analysis_year          # "2025_2026"
AYL <- M$analysis_year_label    # "2026"
if (is.null(M$data_received) || !nzchar(M$data_received)) {
  M$data_received <- "August 25, 2026"
}

## ---- Palette + ggplot theme -----------------------------------------------
COHORT_COL   <- "#3498DB"
BASELINE_COL <- "#E67E22"
LEVEL_COLS   <- c(Low = "#C0392B", Typical = "#95A5A6", High = "#27AE60")
GRID_COL     <- "#D8D8D4"
TEXT_COL     <- "#141410"

theme_ri <- function(base_size = 22) {
  theme_minimal(base_size = base_size, base_family = "Noto Sans") +
    theme(
      text            = element_text(color = TEXT_COL),
      plot.title      = element_text(face = "bold", size = rel(1.05)),
      plot.subtitle   = element_text(color = "#5c5c58"),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = GRID_COL, linewidth = 0.4),
      strip.text      = element_text(face = "bold"),
      legend.position = "top",
      legend.title    = element_blank(),
      axis.title      = element_text(color = "#5c5c58")
    )
}

## ---- Lookups / formatting helpers used inline throughout -------------------
mmn_cell <- function(content, grade_label, norm = c("cohort", "baseline"), year = AY) {
  norm <- match.arg(norm)
  r <- ov[CONTENT_AREA == content & GRADE_LABEL == grade_label & YEAR == year]
  if (!nrow(r) || (norm == "cohort" && r$n_cohort == 0)) return("&mdash;")
  if (norm == "cohort") fmt_mmn(r$mean_cohort, r$median_cohort, r$n_cohort)
  else fmt_mmn(r$mean_baseline, r$median_baseline, r$n_baseline)
}

## Cross-sectional (unconditional) status value for a content area x grade x year
xs_val <- function(content, grade, col = "auc", year = AY) {
  r <- xsec[CONTENT_AREA == content & GRADE == grade & YEAR == year]
  if (!nrow(r)) return(NA_real_)
  r[[col]]
}

## Percent proficient for a content area x grade x year
prof_val <- function(content, grade, year = AY) {
  r <- prof[CONTENT_AREA == content & GRADE == grade & YEAR == year]
  if (!nrow(r)) return(NA_real_)
  r$pct_proficient
}

## mean/median/SD/N cell (SDs added at the committee's request)
fmt_mmsn <- function(mean_val, median_val, sd_val, n) {
  paste0(formatC(round(mean_val, 1), format = "f", digits = 1), "/",
         formatC(round(median_val, 0), format = "d"), "/",
         formatC(round(sd_val, 1), format = "f", digits = 1), "/",
         fmt_count(n))
}

get_val <- function(content, grade_label, col, year = AY) {
  r <- ov[CONTENT_AREA == content & GRADE_LABEL == grade_label & YEAR == year]
  if (!nrow(r)) return(NA_real_)
  r[[col]]
}

## ---- Goodness-of-fit figure helpers (clean vendored SVGs) ------------------
gof_token <- function(content_area) {
  data.table::fcase(
    content_area == "ELA", "ELA",
    content_area == "MATHEMATICS", "MATH",
    content_area == "ELA_SAT", "ELA_SAT",
    content_area == "MATHEMATICS_SAT", "MATH_SAT",
    default = content_area
  )
}

gof_src <- function(content_area, grade, year, norm = "cohort") {
  file.path("Figures", "gof",
            sprintf("%s_%s_%s_g%s.svg", gof_token(content_area), year, norm, grade))
}

## One <figure> (works with the theme figure-modal lightbox: `.reveal figure img`)
gof_figure <- function(content_area, grade, year, norm = "cohort",
                       caption = NULL, alt = NULL, max_height = NULL) {
  src <- gof_src(content_area, grade, year, norm)
  if (!file.exists(src)) {
    return(sprintf('<div style="padding:1em;color:#c0392b;font-size:0.6em">missing: %s</div>', src))
  }
  if (is.null(caption)) caption <- sprintf("%s Grade %s &mdash; %s (%s)",
                                           pretty_subject(content_area), grade,
                                           pretty_year(year), norm)
  if (is.null(alt)) alt <- gsub("&mdash;", "-", caption)
  img_style <- if (!is.null(max_height)) {
    sprintf("max-height:%s;width:auto;max-width:100%%;display:block;margin:0 auto", max_height)
  } else {
    "width:100%;height:auto"
  }
  sprintf(
    '<figure style="margin:0;text-align:center">
       <img src="%s" alt="%s" style="%s"/>
       <figcaption>%s</figcaption>
     </figure>', src, alt, img_style, caption)
}

## Emit raw HTML as a pandoc passthrough block so `native_divs` does not try to
## pair our <div>s with Quarto's auto `::: {.cell}` wrapper (which otherwise
## leaks stray "::" text). Use for every asis chunk that prints raw HTML.
emit_html <- function(x) cat("\n```{=html}\n", x, "\n```\n", sep = "")

## Responsive grid of GoF figures
gof_grid <- function(figs, min_col = "300px", gap = "0.6em", class = NULL) {
  class_attr <- if (!is.null(class)) sprintf(' class="%s"', class) else ""
  emit_html(sprintf(
    '<div%s style="display:grid;grid-template-columns:repeat(auto-fit,minmax(%s,1fr));gap:%s;align-items:start">%s</div>',
    class_attr, min_col, gap, paste(figs, collapse = "")))
}
```

## Roadmap {.center}

::: {.visually-hidden}
Roadmap for the talk
:::

::: {style="font-size: 1.05em;"}
1. **What was run** &mdash; RICAS and PSAT/SAT analyses, cohort and baseline
2. **2026 results** &mdash; overall growth and model fit
3. **Growth in context** &mdash; baseline growth read against the norm period
4. **Grade 11** &mdash; why the baseline results are withdrawn
5. **Growth equity** &mdash; results by student group
6. **Target attainment** &mdash; catch-up / keep-up, and what the target actually is
7. **Questions for the TAC**
:::

::: {.key-insight}
**`r fmt_count(M$n_growth_cohort_total)`** students received a `r AYL` growth percentile across RICAS (grades 4&ndash;8) and SAT (grade 11). Results use the student-level file received **`r M$data_received`**.
:::

::: {.notes}
Revision 2. This version is organised around the committee's written feedback rather than the original results-first order. The substantive change is section 3: baseline growth is now presented next to cross-sectional results from the norm period, which changes how the mathematics trend should be read. Section 4 is new and is the strongest recommendation in the deck. Model-fit slides are confirmation &mdash; flip through unless someone wants a plot enlarged.
:::


## This revision {.smaller}

::: {.key-insight-box}
::: {.key-insight-box-title}
Revision 2 &mdash; revised following Technical Advisory Committee feedback
:::
This deck is a living document. It regenerates from the analysis pipeline, so
feedback that changes an analysis changes every number that depends on it. The
changes below originate in written comments returned by the committee.
:::

::: {.columns}
::: {.column width="49%"}
**What changed**

- Baseline growth is now read **alongside cross-sectional results from the norm period**, following Ho (2009). This changes the reading of the mathematics trend.
- The **grade-11 baseline results are withdrawn** from reporting pending the SAT linking question.
- **Model fit** is reframed as conditional coverage, with the discreteness limit stated explicitly.
- **Catch-up / keep-up** now reports its target horizon, denominators, and intervals.
- **Standard deviations** accompany every group mean.
:::
::: {.column width="49%"}
**What remains open**

- Whether to **re-establish baseline norms**, and on what cadence. The committee's written comments point in both directions; this deck states the trade-off rather than resolving it.
- Whether the ELA growth pattern and the ELA equating pattern are **the same phenomenon**.
- A **form-effects study** using Massachusetts spiralled forms, designed jointly with the validation work.
:::
:::

::: {.notes}
Open by naming the feedback and what it changed. The point of the pipeline is that this took a rebuild, not a rewrite. Two claims in the committee's notes could not be checked against the student-level file and are deliberately not asserted anywhere in this deck: the digital SAT conversion issue, and the form-spiralling contrast between Rhode Island and Massachusetts. Both need confirmation from RIDE, Cognia, or the College Board before they go on a slide.
:::


# SGP results for 2026 {.center}

::: {.visually-hidden}
Section one: 2026 results
:::


## What was run: RICAS (grades 4&ndash;8)

Two SGP analyses were conducted on the `r AYL` RICAS data (ELA and Mathematics):

- **Cohort-referenced** &mdash; growth norms built from the most recent cohorts; by construction these center mean/median SGP on **50**.
- **Baseline-referenced** &mdash; historical coefficient matrices from the pre-pandemic **2017&ndash;18 and 2018&ndash;19** administrations. These are **not** centered on 50 and let us examine COVID-era academic recovery, changes in rates of learning, and possible **scale drift**.
- **Projections** (lagged and straight) were calculated using both norm sets.

::: {.key-insight}
Cohort norms answer *"how did this student grow relative to academic peers this year?"* Baseline norms answer *"how does that growth compare to pre-pandemic rates of learning?"*
:::

::: {.notes}
Same 2017&ndash;18 / 2018&ndash;19 window as SAT. Some skip-year RICAS matrices also use equated 2015&ndash;17 (PARCC-on-RICAS) priors; the most recent baseline year is 2018&ndash;19 throughout. Baseline is the diagnostic lens for recovery, slower learning, and possible drift.
:::


## What was run: PSAT &rarr; SAT (grade 11)

Two SGP analyses were conducted on the PSAT-10 &rarr; SAT progression (single prior):

- **Cohort-referenced** &mdash; centered on 50 by construction.
- **Baseline-referenced** &mdash; the same pre-pandemic **2017&ndash;18 / 2018&ndash;19** window: PSAT-10 (2018) &rarr; SAT (2019).
- **No projections** &mdash; SAT (grade 11) is the last tested grade in Rhode Island.

::: {.theorem-box}
::: {.theorem-box-title}
Single-prior growth
:::
Grade-11 growth uses one prior (PSAT-10). It is a shorter growth window than the RICAS grades, so baseline comparisons are correspondingly more sensitive to form and linking effects.
:::

::: {.notes}
Same 2017&ndash;18 / 2018&ndash;19 baseline window as RICAS; SAT is the single-prior PSAT-10 (2018) to SAT (2019) pair. Flag the shorter window; it matters when we read the SAT baseline numbers.
:::


## Provenance &amp; reproducibility {.smaller}

::: {.columns}
::: {.column width="56%"}
- All analyses used student-level data supplied by the Rhode Island Department of Education, **received `r M$data_received`** (a correction to an earlier file affecting several hundred students).
- Conducted with the **`SGP` R package** (version `r M$sgp_package_version`).
- All analyses converged; goodness-of-fit plots confirm successful analyses.
- This is **revision 2** of the deck, regenerated after TAC feedback; the underlying student-level file is unchanged.
- Source code is public on GitHub:
  - Analyses: [github.com/CenterForAssessment/Rhode_Island](https://github.com/CenterForAssessment/Rhode_Island)
  - SGP package: [github.com/CenterForAssessment/SGP](https://github.com/CenterForAssessment/SGP)
:::
::: {.column width="42%"}
::: {.key-insight-box}
::: {.key-insight-box-title}
Reproducible pipeline
:::
This deck is generated from a cached summary of the student-level LONG data. When RIDE supplies revised data, re-running the pipeline refreshes **every** number and table automatically.
:::
:::
:::

::: {.notes}
This is the August 25 file. At reported precision the grade-level 2026 cells match the earlier run; the correction is incorporated.
:::


## 2026 SGP results at a glance

```{r results-table}
#| output: asis

grades <- c("4", "5", "6", "7", "8", "11 (SAT)")
content_ela  <- function(g) if (g == "11 (SAT)") "ELA_SAT" else "ELA"
content_math <- function(g) if (g == "11 (SAT)") "MATHEMATICS_SAT" else "MATHEMATICS"

rows <- vapply(grades, function(g) {
  sprintf(
    '<tr><th class="rt-grade">%s</th><td>%s</td><td class="rt-base">%s</td><td>%s</td><td class="rt-base">%s</td></tr>',
    g,
    mmn_cell(content_ela(g),  g, "cohort"),
    mmn_cell(content_ela(g),  g, "baseline"),
    mmn_cell(content_math(g), g, "cohort"),
    mmn_cell(content_math(g), g, "baseline")
  )
}, character(1))

emit_html(sprintf('
<style>
  table.results-table { border-collapse: collapse; width: 100%%; font-size: 0.72em; }
  table.results-table th, table.results-table td { padding: 0.35em 0.6em; text-align: center; border-bottom: 1px solid var(--color-border); }
  table.results-table thead .grp { font-family: "Josefin Sans", sans-serif; font-weight: 600; }
  table.results-table thead .grp-ela  { color: %s; border-bottom: 2px solid %s; }
  table.results-table thead .grp-math { color: %s; border-bottom: 2px solid %s; }
  table.results-table .rt-grade { text-align: left; font-family: "Josefin Sans", sans-serif; }
  table.results-table .rt-base { color: %s; }
  table.results-table tbody tr:hover { background: var(--color-surface-subtle); }
  .rt-caption { font-size: 0.55em; color: var(--color-secondary); font-style: italic; margin-top: 0.4em; }
</style>
<table class="results-table">
  <thead>
    <tr>
      <th rowspan="2">Grade</th>
      <th class="grp grp-ela" colspan="2">ELA</th>
      <th class="grp grp-math" colspan="2">Mathematics</th>
    </tr>
    <tr>
      <th>Cohort</th><th class="rt-base">Baseline</th>
      <th>Cohort</th><th class="rt-base">Baseline</th>
    </tr>
  </thead>
  <tbody>%s</tbody>
</table>
<div class="rt-caption">Each cell: mean&#8202;/&#8202;median&#8202;/&#8202;N of student growth percentiles, %s. Cohort norms center on 50 by construction; baseline norms are referenced to pre-pandemic rates of learning. <strong>The grade-11 baseline cells are shown for the committee only</strong> &mdash; this revision recommends withholding them from public reporting (see section 4). Data received %s.</div>
',
COHORT_COL, COHORT_COL, "#7a5c33", "#7a5c33", BASELINE_COL,
paste(rows, collapse = "\n"), AYL, M$data_received))
```

::: {.notes}
Walk the cohort columns first (all ~50, as expected), then the baseline columns &mdash; that is where the story is. The deviations from 50 are what we dig into next. At the reported precision, the August 25 correction does not change these headline cells.
:::


## Reading the 2026 baseline results

```{r baseline-narrative}
#| output: asis

## Flag baseline means that deviate meaningfully from 50 (analysis year, RICAS + SAT).
## RICAS only: the grade-11 baseline results are treated separately (section 4),
## because there the conditional and cross-sectional measurements disagree.
by <- ov[YEAR == AY & n_baseline > 0 & ASSESSMENT == "RICAS"]
by[, dev := mean_baseline - 50]
flag <- by[abs(dev) >= 2][order(-abs(dev))]

mk <- function(x) sprintf("%s Grade %s (mean <strong>%.1f</strong>)",
                          x$SUBJECT, x$GRADE_LABEL, x$mean_baseline)
li <- function(dt) {
  if (!nrow(dt)) return("<li>None materially different</li>")
  paste0("<li>", vapply(seq_len(nrow(dt)), function(i) mk(dt[i]), character(1)), "</li>", collapse = "")
}
above <- flag[dev > 0]; below <- flag[dev < 0]

emit_html(sprintf('
<div style="display:flex;gap:2em;align-items:flex-start">
  <div style="flex:1">
    <p style="font-weight:600;color:#27ae60;margin-bottom:0.2em">Above pre-pandemic rates (baseline mean &gt; 52)</p>
    <ul>%s</ul>
  </div>
  <div style="flex:1">
    <p style="font-weight:600;color:#c0392b;margin-bottom:0.2em">Below pre-pandemic rates (baseline mean &lt; 48)</p>
    <ul>%s</ul>
  </div>
</div>', li(above), li(below)))
```

::: {.key-insight}
Cohort-referenced results are, as expected, centered on 50 with even conditional coverage. The **baseline** results are uneven across grade and content &mdash; the pattern the next section reads against cross-sectional status. Grade 11 is held back for section 4.
:::

::: {.notes}
Generated from the data, so the flagged grades update automatically with a new run. Restricted to RICAS in this revision; grade 11 gets its own section because there the measurements contradict each other. Keep interpretation cautious here and let the context section do the interpreting.
:::


## Growth distribution: Low / Typical / High

```{r level-distribution}
#| fig-width: 16
#| fig-height: 6.2

ld <- lev[YEAR == AY & !is.na(LEVEL)]
ld[, GRADE_LABEL := factor(GRADE_LABEL, levels = c("4","5","6","7","8","11 (SAT)"))]
ld[, NORM := factor(NORM, levels = c("Cohort", "Baseline"))]

ggplot(ld, aes(x = GRADE_LABEL, y = pct, fill = LEVEL)) +
  geom_col(width = 0.78) +
  geom_hline(yintercept = c(33.3, 66.7), color = "white", linewidth = 0.5, linetype = "22") +
  facet_grid(SUBJECT ~ NORM) +
  scale_fill_manual(values = LEVEL_COLS) +
  scale_y_continuous(labels = function(x) paste0(x, "%"), expand = expansion(mult = c(0, 0.02))) +
  labs(x = NULL, y = NULL,
       title = paste0("Share of students at each growth level, ", AYL),
       subtitle = "Cohort norms split roughly one-third each; baseline shifts reveal recovery / drift") +
  theme_ri(20)
```

::: {.notes}
Cohort panels should show the classic ~1/3 Low, Typical, High. Where baseline panels tilt toward Low or High is where pre-pandemic-referenced growth diverges. Point to the largest tilts.
:::


# Model fit and conditional coverage {.center}

::: {.visually-hidden}
Section two: model fit and conditional coverage
:::

::: {.notes}
Renamed in this revision. The committee's objection to the label "goodness of fit" was well taken: the next slide states what these plots do and do not test. Budget ~5 minutes; click-to-enlarge if anyone wants a panel.
:::


## What these plots do and do not diagnose

::: {.tac-tag}
Revised following TAC comment
:::

Each plot shows, for every decile of prior achievement, the share of students falling in each growth-percentile decile. Two different things move those cells, and only one of them is a fitting problem.

::: {.columns}
::: {.column width="48%"}
**What they diagnose**

*Systematic tilt across prior achievement.* If growth percentiles drift upward or downward as prior achievement rises, the conditional model is not covering the range of prior achievement evenly. That is misfit, and it is actionable.
:::
::: {.column width="48%"}
**What they cannot diagnose**

*Departure from exact uniformity.* Scores are discrete. Students sharing a prior and a current score must receive the same growth percentile, so the conditional distribution cannot be uniform and the cells cannot all sit at 10%. That lumpiness follows from fitting a continuous quantile model to a discrete scale.
:::
:::

::: {.key-insight}
The diagnostic content is the **tilt**, not the distance from 10%. The residual lumpiness is a property of the model meeting a discrete scale &mdash; it is the model's constraint, not a defect in the data. For baseline norms a level shift is neither: it is the change in the prior-to-current relationship that the comparison exists to detect.
:::

::: {.notes}
This slide answers a standing TAC objection: calling these "goodness of fit" invited the reading that any departure from uniformity is a failure. It is not, and the discreteness floor belongs to the model rather than to the scale. What remains genuinely diagnostic is conditional tilt, and there is none of consequence here. Say plainly that the label was imprecise and that this is the corrected framing.
:::


## RICAS ELA &mdash; cohort model fit, `r AYL`

```{r gof-ela-cohort}
#| output: asis
gof_grid(vapply(c("4","5","6","7","8"),
                function(g) gof_figure("ELA", g, AY, "cohort",
                                       caption = sprintf("ELA Grade %s", g)),
                character(1)))
```

::: {.notes}
Uniform bands across prior achievement = good fit. All five grades converge cleanly.
:::


## RICAS Mathematics &mdash; cohort model fit, `r AYL`

```{r gof-math-cohort}
#| output: asis
gof_grid(vapply(c("4","5","6","7","8"),
                function(g) gof_figure("MATHEMATICS", g, AY, "cohort",
                                       caption = sprintf("Mathematics Grade %s", g)),
                character(1)))
```


## SAT &mdash; cohort model fit, `r AYL`

```{r gof-sat-cohort}
#| output: asis
gof_grid(c(
  gof_figure("ELA_SAT", "EOCT", AY, "cohort",
             caption = "ELA &mdash; SAT (grade 11)", max_height = "63vh"),
  gof_figure("MATHEMATICS_SAT", "EOCT", AY, "cohort",
             caption = "Mathematics &mdash; SAT (grade 11)", max_height = "63vh")
), min_col = "380px", class = "gof-grid-compact")
```

::: {.notes}
Single-prior SAT fit. Note the shorter growth window relative to RICAS.
:::


## Cohort vs. baseline fit &mdash; ELA Grade 7

```{r gof-contrast}
#| output: asis
gof_grid(c(
  gof_figure("ELA", "7", AY, "cohort",
             caption = "ELA Grade 7 &mdash; Cohort norms", max_height = "63vh"),
  gof_figure("ELA", "7", AY, "baseline",
             caption = "ELA Grade 7 &mdash; Baseline norms", max_height = "63vh")
), min_col = "380px", class = "gof-grid-compact")
```

::: {.key-insight}
Both norm sets fit the data well. The difference is the **reference**: cohort centers on this year's peers; baseline holds the pre-pandemic standard fixed &mdash; so a level shift here is signal, not misfit.
:::


# Growth in context {.center}

::: {.visually-hidden}
Section three: growth in context
:::

::: {.notes}
The analytic heart of the talk (~15 minutes) and the section most changed in this revision. Baseline growth is no longer presented on its own; it is read against cross-sectional status in the norm period, which is what the committee asked for. That pairing reverses the sign of the mathematics story. Slow down here.
:::


## Pandemic-related academic recovery

::: {.columns}
::: {.column width="54%"}
Five years on from the onset of the pandemic, we examine **changes in growth over time** to distinguish among:

- **Ripple effects** &mdash; continued downstream impact for students whose early schooling was disrupted.
- **Slower (or faster) learning gains** &mdash; a real change in how much students learn, relative to 2017&ndash;19.
- **Scale drift / linking** &mdash; movement in the underlying scale that can look like a growth change.
:::
::: {.column width="42%"}
::: {.theorem-box}
::: {.theorem-box-title}
Two views of the same data
:::
**Diagonal** &mdash; follow one cohort as it ages grade-to-grade.

**Vertical** &mdash; hold a grade fixed and watch successive cohorts year-to-year.
:::
:::
:::

::: {.notes}
Set up the two reading directions. Vertical is not a drift test; it is a same-grade comparison of successive cohorts. The following slide on mechanisms is the place to say more.
:::


## Where were they during the pandemic?

```{r pandemic-grid}
#| output: asis
## Grade in 2026 -> grade during the two pandemic school years.
tbl <- data.frame(
  cur   = c("Grade 4","Grade 5","Grade 6","Grade 7","Grade 8","Grade 11 (SAT)"),
  s2020 = c("PK","K","Grade 1","Grade 2","Grade 3","Grade 6"),
  s2021 = c("K","Grade 1","Grade 2","Grade 3","Grade 4","Grade 7")
)
rows <- apply(tbl, 1, function(r)
  sprintf('<tr><th style="text-align:left;font-family:\'Josefin Sans\',sans-serif">%s</th><td>%s</td><td>%s</td></tr>',
          r["cur"], r["s2020"], r["s2021"]))
emit_html(sprintf('
<table class="results-table" style="font-size:0.7em">
  <thead><tr><th>Current (2026)</th><th>Spring 2020</th><th>2020&ndash;2021</th></tr></thead>
  <tbody>%s</tbody>
</table>
<div class="rt-caption">The lower grades tested in 2026 spent their earliest, most foundational schooling years during pandemic disruption.</div>',
paste(rows, collapse = "")))
```

::: {.notes}
The grade-7 and grade-8 cohorts (early-literacy grades during the pandemic) are the ones we track next.
:::


## Trend: same grade over time (drift indicator)

```{r trend-samegrade}
#| fig-width: 16
#| fig-height: 6.2

tr <- ov[ASSESSMENT == "RICAS" & n_baseline > 0 &
         YEAR %in% c("2022_2023","2023_2024","2024_2025","2025_2026")]
tr[, GRADE := factor(GRADE, levels = as.character(4:8))]

ggplot(tr, aes(x = YEAR_LABEL, y = mean_baseline, color = GRADE, group = GRADE)) +
  geom_hline(yintercept = 50, color = "#9a9a94", linetype = "22") +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.6) +
  facet_wrap(~ SUBJECT) +
  scale_color_brewer(palette = "Dark2", name = "Grade") +
  labs(x = NULL, y = "Mean baseline SGP",
       title = "Baseline-referenced growth by grade, over time",
       subtitle = "Same grade, successive cohorts, baseline-referenced. Read on its own this looks like a mathematics problem. The next slides show why it is not.") +
  theme_ri(20) + theme(legend.position = "right")
```

::: {.notes}
Vertical reading: same grade, successive cohorts. Mathematics grades 4&ndash;6 show the multi-year decline. In the previous revision this was where the deck raised the alarm. Do not do that now &mdash; set it up as the puzzle and go to the context slides, which resolve most of it.
:::


## Baseline growth needs a second measurement

::: {.tac-tag}
Added following TAC comment
:::

A mean baseline SGP is a **conditional** statement: given where these students started, how does their current standing compare with the 2017&ndash;19 relationship between prior and current scores? It cannot, alone, separate a change in learning from a change in the scale.

The committee's remedy is to put it beside the **cross-sectional** comparison from the same norm period &mdash; a distributional comparison that assumes no growth model at all, following Ho (2009).

::: {.theorem-box}
::: {.theorem-box-title}
The two are the same statistic, with and without conditioning
:::
Drop the prior-score conditioning from a baseline SGP and it becomes the percentile of a student's score in the norm-period **marginal** distribution. Averaged over the cohort, that is

$$E\left[F_b(Y_c)\right] = \Pr(Y_b < Y_c) = \mathrm{AUC}, \qquad V = \sqrt{2}\,\Phi^{-1}(\mathrm{AUC})$$

the Mann&ndash;Whitney statistic behind Ho's scale-invariant $V$. The cross-sectional comparison is the **zero-covariate case** of the baseline SGP &mdash; not a rival method.
:::

::: {.notes}
This is the conceptual pivot of the revision, and it is worth being precise about the credit. The committee proposed CDF-based comparisons as a better drift diagnostic than SGPs. The relationship is tighter than either framing suggests: they are the same estimand with and without conditioning on prior achievement. Conditioning buys removal of the largest source of cohort composition change; it costs a model, and it inherits any drift in the PRIOR year's scale. The unconditional version assumes nothing and inherits nothing, but confounds composition with scale with learning. Neither identifies drift alone. Reference: Ho, A. D. (2009), A nonparametric framework for comparing trends and gaps across tests, JEBS 34, 201-228.
:::


## One statistic, with and without conditioning

```{r cdf-identity-figure}
#| output: asis
emit_html('<figure style="margin:0 0 0.15em 0;text-align:center">
  <img src="Figures/sgp_cdf_identity.svg"
       alt="Three panels. A: the norm-period distribution of the current score at each prior score, with the lower tail below one student\'s score shaded - the baseline SGP. B: those conditional distributions averaged over the prior into a single marginal distribution, with the area below the same student\'s score shaded. C: the probability-probability plot, with the area above the curve shaded and labelled AUC. Below: the expectation of F_b(Y_c) equals the probability that Y_b is less than Y_c, which equals AUC, and V equals root two times the inverse normal of AUC."
       style="width:100%;height:auto"/>
</figure>')
```

::: {.key-insight}
**A** is the baseline SGP. **B** is the same quantity with the prior dropped. **C** averages **B** over the cohort &mdash; and lands exactly on the Mann&ndash;Whitney statistic behind Ho's $V$. The cross-sectional comparison is not a rival to the baseline SGP; it is the **zero-covariate case** of it.
:::

::: {.notes}
Walk the three panels rather than the algebra. Panel A: a baseline SGP asks where a student sits in the norm-period distribution of students who started where they started. Panel B: erase the prior and the family of conditional distributions collapses into one marginal; the same student now has an unconditional percentile. Panel C: average that over the cohort and you have the probability that a randomly drawn current-year student outscores a randomly drawn norm-period student, which is Ho's statistic, and geometrically it is the area above the PP curve on the previous slide.

Both halves of this were checked against the 2026 data rather than asserted. For ELA grade 6 the mean unconditional baseline percentile is 43.01 and the AUC computed from the two distributions is 43.01. The area above the fitted PP curve reproduces the AUC in all ten RICAS grade-by-subject cells to a tenth of a point.

The payoff is the division of labour. Conditioning buys removal of the largest single source of cohort composition change, prior achievement; it costs a quantile-regression model and it inherits any drift in the PRIOR year's scale. Dropping the conditioning assumes nothing and inherits nothing, but confounds composition with scale with learning. Neither identifies drift on its own, which is exactly why this deck now reports both.
:::


## The same grades, measured both ways

```{r context-joint}
#| fig-width: 16
#| fig-height: 6.4

yrs <- c("2021_2022", "2022_2023", "2023_2024", "2024_2025", "2025_2026")
gr5 <- as.character(4:8)

cond <- ov[ASSESSMENT == "RICAS" & GRADE %in% gr5 & YEAR %in% yrs & n_baseline > 0,
           .(SUBJECT, GRADE, YEAR_LABEL, value = mean_baseline)]
cond[, measure := "Conditional — mean baseline SGP"]

unc <- xsec[ASSESSMENT == "RICAS" & GRADE %in% gr5 & YEAR %in% yrs,
            .(SUBJECT, GRADE, YEAR_LABEL, value = auc)]
unc[, measure := paste0("Unconditional — status vs ", BNYL)]

cmb <- rbind(cond, unc)
cmb[, GRADE := factor(GRADE, levels = gr5)]
cmb[, measure := factor(measure, levels = c(paste0("Unconditional — status vs ", BNYL),
                                            "Conditional — mean baseline SGP"))]

ggplot(cmb, aes(x = YEAR_LABEL, y = value, color = GRADE, group = GRADE)) +
  geom_hline(yintercept = 50, color = "#9a9a94", linetype = "22") +
  geom_line(linewidth = 1.05) +
  geom_point(size = 2.4) +
  facet_grid(SUBJECT ~ measure) +
  scale_color_brewer(palette = "Dark2", name = "Grade") +
  labs(x = NULL, y = "Percentile points (50 = norm period)",
       title = "Conditional growth and unconditional status, against the same 2017–19 reference",
       subtitle = "Left: where this year's distribution sits in the norm-period distribution. Right: mean baseline SGP.") +
  theme_ri(19) + theme(legend.position = "right")
```

::: {.notes}
The two columns answer different questions against the same reference. Mathematics: status climbs every year in every grade and is now at or above the norm period, while baseline SGP falls toward 50 from well above it. ELA: status is flat and below the norm period throughout, and baseline SGP sits below 50 in most grades. Let the audience read the divergence before you interpret it on the next slide.
:::


## Reading the two together

::: {.columns}
::: {.column width="49%"}
**Mathematics &mdash; recovery completing**

Status rose in every grade every year: grade 4 from `r sprintf("%.0f", xs_val("MATHEMATICS","4",year="2021_2022"))` to `r sprintf("%.0f", xs_val("MATHEMATICS","4"))`, grade 8 from `r sprintf("%.0f", xs_val("MATHEMATICS","8",year="2021_2022"))` to `r sprintf("%.0f", xs_val("MATHEMATICS","8"))`.

The falling baseline SGP is the **fading of rebound growth off a depressed base**, not deterioration. Growth far above the pre-pandemic rate was what recovery required; arriving back at 50 is what its completion looks like.
:::
::: {.column width="49%"}
**ELA &mdash; the genuine concern**

Status is flat and below the norm period in every grade and every year (`r sprintf("%.0f", xs_val("ELA","5"))` at grade 5, `r sprintf("%.0f", xs_val("ELA","6"))` at grade 6), and baseline SGP is below 50 in the same grades.

Both measurements agree, and neither is recovering. This, not the mathematics line, is where the deck's concern belongs.
:::
:::

::: {.key-insight}
The measurements move **in opposite directions across subjects**. A program-wide linking or equating event would move grades and subjects together in one year. That is not what the RICAS data show &mdash; which argues against a common scale-drift explanation and for reading these as real, subject-specific changes.
:::

::: {.notes}
This slide reverses the previous revision's emphasis, and say so out loud. Last version led with the mathematics decline as the worry and held the drift question open. Adding the cross-sectional context resolves most of it: mathematics status is up, so the falling baseline SGP is recovery completing rather than slowing learning. ELA is down on both measures and is the finding that deserves attention. The one grade that stands out, ELA grade 7, is high on both measures in every year 2022 through 2026, not just this year, so a 2026 equating event is an unlikely explanation for it.
:::


## The distributional comparison itself

```{r pp-plot}
#| fig-width: 16
#| fig-height: 5.9

pl <- ppc[ASSESSMENT == "RICAS" & GRADE %in% as.character(4:8)]
pl[, GRADE := factor(paste("Grade", GRADE), levels = paste("Grade", 4:8))]

ggplot(pl, aes(x = p_base, y = p_cur)) +
  geom_abline(slope = 1, intercept = 0, color = "#9a9a94", linetype = "22") +
  geom_line(aes(color = SUBJECT), linewidth = 1.05) +
  facet_grid(SUBJECT ~ GRADE) +
  scale_color_manual(values = c(ELA = COHORT_COL, Mathematics = "#7a5c33"), guide = "none") +
  scale_x_continuous(breaks = c(0, 50, 100)) +
  scale_y_continuous(breaks = c(0, 50, 100)) +
  coord_fixed() +
  labs(x = paste0("Percentile in the ", BNYL, " distribution"),
       y = paste0(AYL, " percentile"),
       title = paste0("Probability–probability plots: ", AYL, " against ", BNYL),
       subtitle = "Above the diagonal = more students below that score than in the norm period, i.e. the distribution has moved down. Invariant to any monotone rescaling.") +
  theme_ri(17)
```

::: {.key-insight}
ELA sits above the diagonal at every grade and across the whole range &mdash; a uniform downward shift, not a tail effect. Mathematics tracks the diagonal closely, with grades 4 and 5 slightly below it.
:::

::: {.notes}
This is the display the committee asked for, and its value is that it makes no distributional or model assumption and survives any monotone rescaling of the reported scale. A curve that departs from the diagonal only at one end would point to a tail or ceiling problem; a parallel offset across the range, which is what ELA shows, is a whole-distribution shift. Note the visible stepping: it is the discreteness of the reported scale, the same discreteness discussed in the model-fit section.
:::


## Do the two measurements agree?

```{r prof-scatter}
#| fig-width: 15
#| fig-height: 6.0

p19 <- prof[YEAR == BNY & ASSESSMENT == "RICAS", .(CONTENT_AREA, GRADE, prof_base = pct_proficient)]
p26 <- prof[YEAR == AY  & ASSESSMENT == "RICAS", .(CONTENT_AREA, GRADE, SUBJECT, prof_now = pct_proficient)]
sc  <- merge(p26, p19, by = c("CONTENT_AREA", "GRADE"))
sc  <- merge(sc, ov[YEAR == AY & ASSESSMENT == "RICAS" & n_baseline > 0,
                    .(CONTENT_AREA, GRADE, mean_baseline)],
             by = c("CONTENT_AREA", "GRADE"))
sc[, prof_chg := prof_now - prof_base]

ggplot(sc, aes(x = prof_chg, y = mean_baseline, color = SUBJECT)) +
  geom_hline(yintercept = 50, color = "#9a9a94", linetype = "22") +
  geom_vline(xintercept = 0, color = "#9a9a94", linetype = "22") +
  geom_point(size = 4.2) +
  geom_text(aes(label = GRADE), vjust = -1.15, size = 5.5, show.legend = FALSE) +
  scale_color_manual(values = c(ELA = COHORT_COL, Mathematics = "#7a5c33")) +
  scale_x_continuous(expand = expansion(mult = 0.12)) +
  scale_y_continuous(expand = expansion(mult = 0.12)) +
  labs(x = paste0("Change in % proficient since ", BNYL, " (percentage points)"),
       y = "Mean baseline SGP",
       title = "Naive status trend against baseline-referenced growth",
       subtitle = "Points are grades. Agreement between the two suggests a real trend; disagreement is the diagnostic signature worth investigating.") +
  theme_ri(19)
```

::: {.key-insight}
**ELA agrees strongly across the two measurements** &mdash; grade 7 highest on both, grades 5 and 6 lowest on both. **Mathematics sits close to the reference on both**: status slightly above `r BNYL`, conditional growth slightly below it. That combination describes a cohort arriving in the grade already ahead of its `r BNYL` counterpart, not a scale artifact &mdash; which would show up as a grade far from the reference on one measure and unmoved on the other.
:::

::: {.notes}
This is the twelve-point plot the committee asked for two meetings running, now standing. The criterion is the committee's own: where the naive trend and the growth measure agree, the trend is probably real; where a grade shows flat status and a baseline SGP well below 50, or the reverse, that is where to look. Nothing here is badly off-diagonal for RICAS. Grade 11 is deliberately absent and is the subject of the next section, because it is precisely the case where the two measurements contradict each other.
:::


## ELA Grade 7 over time &mdash; goodness of fit

```{r gof-drift-ela7}
#| output: asis
gof_grid(vapply(c("2022_2023","2023_2024","2024_2025","2025_2026"),
                function(y) gof_figure("ELA", "7", y, "baseline",
                                       caption = sprintf("ELA Grade 7 &mdash; %s", pretty_year(y))),
                character(1)), min_col = "300px")
```

::: {.notes}
Same grade, four administrations, baseline norms. Consistent shape = the quantile regression still fits; a level shift is a change in the rate of learning, not misfit. Advance unless someone wants a panel enlarged.
:::


## Mathematics Grade 6 over time &mdash; goodness of fit

```{r gof-drift-math6}
#| output: asis
gof_grid(vapply(c("2022_2023","2023_2024","2024_2025","2025_2026"),
                function(y) gof_figure("MATHEMATICS", "6", y, "baseline",
                                       caption = sprintf("Math Grade 6 &mdash; %s", pretty_year(y))),
                character(1)), min_col = "300px")
```


## Diagonal: following the 2026 Grade-8 cohort

```{r gof-diagonal-ela8}
#| output: asis
gof_grid(c(
  gof_figure("ELA", "6", "2023_2024", "baseline", caption = "Grade 6 &mdash; 2024"),
  gof_figure("ELA", "7", "2024_2025", "baseline", caption = "Grade 7 &mdash; 2025"),
  gof_figure("ELA", "8", "2025_2026", "baseline", caption = "Grade 8 &mdash; 2026")
), min_col = "340px")
```

::: {.key-insight}
The same students, three consecutive years (ELA). This cohort was in the early-literacy grades during the pandemic &mdash; the diagonal lets us watch their baseline-referenced growth as they progressed.
:::

::: {.notes}
Diagonal reading: this is one cohort aging through the system, not a fixed grade. Contrast with the previous drift slides.
:::


# Grade 11: a linking question {.center}

::: {.visually-hidden}
Section four: grade 11
:::

::: {.notes}
New in this revision, and the strongest recommendation in the deck. Budget ~6 minutes and do not rush the recommendation slide.
:::


## Two measurements that cannot both be right

::: {.columns}
::: {.column width="48%"}
**What the baseline SGPs say**

- ELA: mean/median **`r sprintf("%.1f", get_val("ELA_SAT","11 (SAT)","mean_baseline"))` / `r sprintf("%.0f", get_val("ELA_SAT","11 (SAT)","median_baseline"))`**
- Mathematics: **`r sprintf("%.1f", get_val("MATHEMATICS_SAT","11 (SAT)","mean_baseline"))` / `r sprintf("%.0f", get_val("MATHEMATICS_SAT","11 (SAT)","median_baseline"))`**

Read literally: the typical Rhode Island eleventh grader grew at roughly the **`r sprintf("%.0f", get_val("ELA_SAT","11 (SAT)","median_baseline"))`th percentile** of the pre-pandemic growth distribution in ELA.
:::
::: {.column width="48%"}
**What every other measurement says**

- ELA status vs `r BNYL`: **`r sprintf("%.0f", xs_val("ELA_SAT","EOCT"))`** (50 = unchanged)
- Mathematics status vs `r BNYL`: **`r sprintf("%.0f", xs_val("MATHEMATICS_SAT","EOCT"))`**
- Percent proficient, ELA: `r sprintf("%.1f%%", prof_val("ELA_SAT","EOCT", BNY))` &rarr; **`r sprintf("%.1f%%", prof_val("ELA_SAT","EOCT"))`**
- Percent proficient, mathematics: `r sprintf("%.1f%%", prof_val("MATHEMATICS_SAT","EOCT", BNY))` &rarr; **`r sprintf("%.1f%%", prof_val("MATHEMATICS_SAT","EOCT"))`**

Grade-11 achievement is **down** on every cross-sectional measure.
:::
:::

::: {.key-insight}
Exceptional growth against a pre-pandemic reference, in a cohort whose measured achievement has fallen relative to that same reference, is not a coherent description of Rhode Island eleventh graders. The most likely explanation is the **PSAT-10 to SAT relationship**, not the students.
:::

::: {.notes}
This is the committee's triangulation principle applied to the case where it bites hardest. Note what changed between the norm period and now: the 2018 and 2019 baseline is the paper SAT, the 2026 administration is digital and adaptive. A change in the prior-to-current relationship of exactly that kind would inflate conditional growth while leaving status alone or depressed, which is what we see. Be careful to claim only what the data support: the contradiction is established here; the specific mechanism is not.
:::


## The shape of the grade-11 distribution

```{r sat-distribution}
#| fig-width: 16
#| fig-height: 5.8

d11 <- sgpd[ASSESSMENT == "SAT"]
d11[, dec_label := factor(DECILE, levels = 0:9,
      labels = sprintf("%d–%d", (0:9) * 10 + 1, pmin(99L, ((0:9) + 1L) * 10L)))]
d11[, NORM := factor(NORM, levels = c("Cohort", "Baseline"))]

ggplot(d11, aes(x = dec_label, y = pct, fill = NORM)) +
  geom_hline(yintercept = 10, color = "#9a9a94", linetype = "22") +
  geom_col(width = 0.8) +
  facet_grid(SUBJECT ~ NORM) +
  scale_fill_manual(values = c(Cohort = COHORT_COL, Baseline = BASELINE_COL), guide = "none") +
  scale_y_continuous(labels = function(x) paste0(x, "%"),
                     expand = expansion(mult = c(0, 0.05))) +
  labs(x = "Growth percentile", y = NULL,
       title = paste0("Full grade-11 growth percentile distributions, ", AYL),
       subtitle = "Dashed line = the 10% each decile would hold in the reference population") +
  theme_ri(18) + theme(axis.text.x = element_text(size = rel(0.72)))
```

::: {.key-insight}
Cohort norms are flat, as they must be. The ELA **baseline** distribution rises monotonically to **`r sprintf("%.0f%%", sgpd[CONTENT_AREA=="ELA_SAT" & NORM=="Baseline" & DECILE==9, pct])`** in the top decile &mdash; nearly three times its expected share. A shape like that is a property of the comparison, not of a cohort.
:::

::: {.notes}
The committee asked to see the distribution rather than the summary, on the reasoning that growth percentiles are built to be roughly uniform in the reference population, so a strongly skewed baseline distribution is itself informative. It was the right request. The gap between the ELA baseline mean and median follows directly from this pile-up at the ceiling.
:::


## Recommendation on grade-11 baseline reporting

::: {.warning-box}
::: {.warning-box-title}
Withhold grade-11 baseline SGPs from public reporting
:::
Grade-11 **baseline** growth percentiles should be withheld from public reporting until the PSAT-10 to SAT linking question is resolved, and the reason should be stated publicly rather than left as a silent omission. Grade-11 **cohort** growth percentiles are unaffected and remain suitable for reporting and accountability use.
:::

::: {.columns}
::: {.column width="49%"}
**Why now**

- The conditional and cross-sectional measurements contradict each other.
- The baseline distribution is piled at its ceiling.
- The growth window is a single prior, so the estimate is maximally sensitive to any change in the prior-to-current relationship.
:::
::: {.column width="49%"}
**What would resolve it**

- Confirmation from the College Board on the comparability of the current score conversion with the `r BNYL` administration.
- Re-estimation of the grade-11 baseline against a post-transition norm period once one exists.
- Until then, report grade-11 growth on cohort norms only.
:::
:::

::: {.notes}
Say plainly that this is a change from the previous revision, which reported these numbers with a caveat. A caveat is not sufficient when the headline figure is a median growth percentile in the seventies that we do not believe. Note also what this deck deliberately does not assert: no claim is made here about any specific defect in the digital conversion, because that could not be verified from the student-level file. The case for withholding rests entirely on the internal contradiction, which is verifiable.
:::


# Growth equity {.center}

::: {.visually-hidden}
Section five: growth by student group
:::

::: {.notes}
Descriptive equity lens: are growth rates comparable across groups? Keep it factual.
:::


## Growth by student group (RICAS, `r AYL`)

```{r equity-table}
#| output: asis

dims <- c("English Learner", "Students w/ Disability", "Economically Disadv.")
sp <- sgp_pool[ASSESSMENT == "RICAS" & DIMENSION %in% dims]

cell <- function(subj, grp, col) {
  r <- sp[SUBJECT == subj & GROUP == grp]
  if (!nrow(r)) return("&mdash;")
  fmt_mmsn(r[[paste0("mean_", col)]], r[[paste0("median_", col)]],
           r[[paste0("sd_", col)]], r[[paste0("n_", col)]])
}

body <- ""
for (d in dims) {
  grps <- sp[DIMENSION == d, unique(GROUP)]
  body <- paste0(body, sprintf('<tr class="dim-row"><th colspan="5">%s</th></tr>', d))
  for (grp in grps) {
    body <- paste0(body, sprintf(
      '<tr><th class="rt-grade">%s</th><td>%s</td><td class="rt-base">%s</td><td>%s</td><td class="rt-base">%s</td></tr>',
      grp, cell("ELA", grp, "cohort"), cell("ELA", grp, "baseline"),
      cell("Mathematics", grp, "cohort"), cell("Mathematics", grp, "baseline")))
  }
}

emit_html(sprintf('
<style>
  table.results-table tr.dim-row th { text-align:left; background: var(--color-surface-subtle);
    font-family:"Josefin Sans",sans-serif; color: var(--color-primary); padding-top:0.5em; }
</style>
<table class="results-table" style="font-size:0.62em">
  <thead>
    <tr><th rowspan="2">Group</th>
        <th class="grp grp-ela" colspan="2">ELA</th>
        <th class="grp grp-math" colspan="2">Mathematics</th></tr>
    <tr><th>Cohort</th><th class="rt-base">Baseline</th><th>Cohort</th><th class="rt-base">Baseline</th></tr>
  </thead>
  <tbody>%s</tbody>
</table>
<div class="rt-caption">mean&#8202;/&#8202;median&#8202;/&#8202;SD&#8202;/&#8202;N, RICAS grades 4&ndash;8 pooled, %s. Standard deviations added at the committee\'s request.</div>', body, AYL))
```

::: {.notes}
Standard deviations are new this revision. The committee expected the spread to carry the equity story; it does not. Every group sits near 29, within a point of every other group, so the differences are entirely in central tendency. Worth saying explicitly, because it rules out a real possibility rather than merely confirming the obvious.
:::


## Equity gaps at a glance

```{r equity-plot}
#| fig-width: 16
#| fig-height: 6.0

dims <- c("English Learner", "Students w/ Disability", "Economically Disadv.")
ep <- sgp_pool[ASSESSMENT == "RICAS" & DIMENSION %in% dims,
               .(mean_sgp = weighted.mean(mean_cohort, n_cohort),
                 mean_base = weighted.mean(mean_baseline, n_baseline)),
               by = .(DIMENSION, GROUP)]
epl <- melt(ep, id.vars = c("DIMENSION","GROUP"),
            measure.vars = c("mean_sgp","mean_base"),
            variable.name = "norm", value.name = "sgp")
epl[, norm := factor(norm, labels = c("Cohort", "Baseline"))]
epl[, GROUP := gsub(" \\(.*\\)$", "", GROUP)]

ggplot(epl, aes(x = sgp, y = reorder(GROUP, sgp), fill = norm)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.65) +
  geom_vline(xintercept = 50, color = "#9a9a94", linetype = "22") +
  facet_wrap(~ DIMENSION, scales = "free_y", ncol = 1) +
  scale_fill_manual(values = c(Cohort = COHORT_COL, Baseline = BASELINE_COL)) +
  labs(x = "Mean SGP (RICAS ELA + Math pooled)", y = NULL,
       title = paste0("Growth by student group, ", AYL)) +
  theme_ri(19)
```

::: {.notes}
The dashed line is 50. Cohort bars near 50 = comparable current-year growth; the cohort-to-baseline distance shows the pre-pandemic-referenced gap. Descriptive, not causal.
:::


## Within-year growth is close to comparable across groups

::: {.tac-tag}
Elevated following TAC comment
:::

```{r equity-comparable}
#| output: asis

pairs <- list(
  c("English Language Learners (ELL)", "Non-English Language Learners (non-ELL)"),
  c("Students with Disabilities (IEP)", "Students without Disabilities (Non-IEP)"),
  c("Economically Disadvantaged", "Not Economically Disadvantaged")
)
labs_ <- c("English learners", "Students with disabilities", "Economically disadvantaged")

mk <- function(pr, lab) {
  vals <- vapply(c("ELA", "Mathematics"), function(sj) {
    a <- sgp_pool[ASSESSMENT == "RICAS" & SUBJECT == sj & GROUP == pr[1], mean_cohort]
    b <- sgp_pool[ASSESSMENT == "RICAS" & SUBJECT == sj & GROUP == pr[2], mean_cohort]
    sprintf("%.1f vs %.1f", a, b)
  }, character(1))
  sprintf('<tr><th class="rt-grade">%s</th><td>%s</td><td>%s</td></tr>', lab, vals[1], vals[2])
}

emit_html(sprintf('
<table class="results-table" style="font-size:0.68em;max-width:60%%;margin:0 auto">
  <thead><tr><th>Group vs. comparison group</th><th>ELA</th><th>Mathematics</th></tr></thead>
  <tbody>%s</tbody>
</table>
<div class="rt-caption">Mean cohort-referenced SGP, RICAS grades 4&ndash;8 pooled, %s. Standard deviations differ by less than one point across every one of these groups.</div>',
paste(vapply(seq_along(pairs), function(i) mk(pairs[[i]], labs_[i]), character(1)), collapse = ""), AYL))
```

::: {.key-insight}
Where achievement gaps are large, **within-year growth is not**. Cohort SGPs sit within a few points of 50 for every group. Students with disabilities are the one substantial exception, and that gap is consistent across both subjects.
:::

::: {.notes}
The committee's point, and it is well taken: this is a more actionable finding than anything in the baseline columns, and the previous revision buried it in a table. Comparable growth alongside large achievement gaps means the gaps are being maintained rather than widened, which locates the problem in accumulated prior learning rather than in current instruction. The disability gap is the exception and should be named as the priority. Descriptive, not causal.
:::


# Target attainment {.center}

::: {.visually-hidden}
Section six: catch-up / keep-up
:::


## What the target actually is

::: {.tac-tag}
Revised following TAC comment
:::

"Catch up" and "keep up" mean nothing without a stated horizon. In Rhode Island the target is proficiency **by the end of the RICAS progression**, so the horizon is not constant &mdash; it shrinks as students approach grade 8.

```{r targets-horizon}
#| output: asis

hz <- tg[METRIC == "Catch Up / Keep Up" & NORM == "Cohort" & ASSESSMENT == "RICAS",
         .(n = sum(n), attained = sum(attained),
           yrs = max(years_to_target, na.rm = TRUE)),
         by = .(GRADE_LABEL, aspiration)]
hz[, rate := 100 * attained / n]

grades_h <- as.character(4:8)
rows_h <- vapply(grades_h, function(gl) {
  cu <- hz[GRADE_LABEL == gl & aspiration == "Catch Up"]
  ku <- hz[GRADE_LABEL == gl & aspiration == "Keep Up"]
  sprintf('<tr><th class="rt-grade">Grade %s</th><td><strong>%s</strong></td><td>%.1f%%</td><td>%.1f%%</td></tr>',
          gl,
          if (nrow(cu)) ifelse(cu$yrs == 0, "same year", sprintf("%d year%s", as.integer(cu$yrs),
                                ifelse(cu$yrs == 1, "", "s"))) else "&mdash;",
          if (nrow(cu)) cu$rate else NA_real_,
          if (nrow(ku)) ku$rate else NA_real_)
}, character(1))

emit_html(sprintf('
<table class="results-table" style="font-size:0.66em;max-width:70%%;margin:0 auto">
  <thead><tr><th>Grade</th><th>Horizon to target</th><th>Catch-up rate</th><th>Keep-up rate</th></tr></thead>
  <tbody>%s</tbody>
</table>
<div class="rt-caption">Cohort targets, RICAS, ELA and mathematics pooled, %s. Horizon is the median number of years to the target.</div>',
paste(rows_h, collapse = ""), AYL))
```

::: {.key-insight}
As the horizon shrinks from three years to none, catch-up rates fall and keep-up rates rise &mdash; monotonically, in both subjects. Most of the apparent grade gradient is the **target**, not the students. Grade-to-grade comparisons are not meaningful unless the horizon travels with them.
:::

::: {.notes}
The committee asked that the target be made explicit, on the grounds that leaving it out effectively changes the standard across grades. Checking it in the data showed the problem is larger than the comment implied: the horizon runs from three years at grade 4 to zero at grade 8, where "catch up" collapses into "is proficient now" and "keep up" into "is still proficient now". That is why the grade-8 catch-up rate is the lowest and the grade-8 keep-up rate the highest. The committee's further suggestion — that these labels earn their keep in individual student reporting rather than aggregate monitoring — is well taken, and one aggregate figure is retained here only because it is already in use.
:::


## Catch-up / keep-up attainment (RICAS, `r AYL`)

```{r targets-plot}
#| fig-width: 16
#| fig-height: 5.9

ct <- tg[METRIC == "Catch Up / Keep Up" & NORM == "Cohort" & ASSESSMENT == "RICAS"]
ct[, GRADE_LABEL := factor(GRADE_LABEL, levels = as.character(4:8))]
ct[, aspiration := factor(aspiration, levels = c("Catch Up", "Keep Up"))]
ct[, xlab := paste0(GRADE_LABEL, "\n(", ifelse(years_to_target == 0, "same yr",
                    paste0(years_to_target, " yr")), ")")]
ct[, xlab := factor(xlab, levels = unique(xlab[order(GRADE_LABEL)]))]

ggplot(ct, aes(x = xlab, y = rate, fill = aspiration)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.68) +
  geom_errorbar(aes(ymin = rate_lo, ymax = rate_hi),
                position = position_dodge(width = 0.75), width = 0.16,
                linewidth = 0.5, color = "#141410") +
  geom_text(aes(label = fmt_count(n), y = 3),
            position = position_dodge(width = 0.75), size = 3.6, color = "white") +
  facet_wrap(~ SUBJECT) +
  scale_fill_manual(values = c("Catch Up" = "#8E44AD", "Keep Up" = COHORT_COL)) +
  scale_y_continuous(labels = function(x) paste0(x, "%"),
                     limits = c(0, 100), expand = expansion(mult = c(0, 0.02))) +
  labs(x = "Grade (horizon to target)", y = "% attaining target",
       title = "Share of students on track to catch up / keep up",
       subtitle = "Catch Up = students below proficient; Keep Up = students at/above. Bars show 95% Wilson intervals; N is printed in each bar.") +
  theme_ri(19)
```

::: {.notes}
Same figure as the previous revision with three additions the committee asked for: the horizon on the axis, denominators in the bars, and intervals. Read across grades only with the horizon in view. The intervals are narrow because the denominators are large; they are here so the rates are falsifiable, not because precision is in doubt.
:::


## Target attainment by prior achievement

```{r targets-highneed}
#| output: asis

th <- tgh[NORM == "Cohort"]
low  <- th[grepl("< 25", HIGH_NEED_STATUS) & aspiration == "Catch Up"]
high <- th[grepl("> 75", HIGH_NEED_STATUS) & aspiration == "Keep Up"]

emit_html(sprintf('
<div style="display:flex;gap:2em;justify-content:center;margin-top:0.5em">
  <div class="stat-callout"><div class="stat-number">%.0f%%</div>
    <div class="stat-label">of lowest-prior-achievement students (&lt;25th %%ile)<br/>are on track to <strong>catch up</strong><br/>
      <span style="font-size:0.82em;opacity:0.75">N = %s &middot; 95%% CI %.1f&ndash;%.1f%%</span></div></div>
  <div class="stat-callout"><div class="stat-number">%.0f%%</div>
    <div class="stat-label">of highest-prior-achievement students (&gt;75th %%ile)<br/>are on track to <strong>keep up</strong><br/>
      <span style="font-size:0.82em;opacity:0.75">N = %s &middot; 95%% CI %.1f&ndash;%.1f%%</span></div></div>
</div>
<div class="rt-caption" style="text-align:center">These are different students judged against different targets, so the two figures are not a gap. Denominators and intervals added at the committee\'s request.</div>',
low$rate, fmt_count(low$n), low$rate_lo, low$rate_hi,
high$rate, fmt_count(high$n), high$rate_lo, high$rate_hi))
```

::: {.key-insight}
Students entering below the 25th percentile face the steepest catch-up targets and attain them least often &mdash; the central equity concern for accountability use of these growth results.
:::

::: {.notes}
Denominators and intervals are new. Say the caveat out loud: these two percentages describe different students measured against different targets, so quoting them as a gap would be wrong. What the pair does establish is that the catch-up target is far harder to reach than the keep-up target, and that the students facing it are the ones furthest behind.
:::


## Summary {.smaller}

- **Cohort-referenced SGPs** show even conditional coverage across RICAS and SAT &mdash; suitable for accountability calculations.
- **Mathematics has recovered.** Status is back to or above the `r BNYL` distribution in every grade. The falling baseline SGP is rebound growth fading as recovery completes, not a decline in learning. This reverses the previous revision's reading.
- **ELA is the genuine concern.** It sits below the norm period on both the conditional and the unconditional measure, in most grades, and is not recovering.
- **A common scale-drift explanation is not supported for RICAS.** The two subjects move in opposite directions; a linking event would move grades and subjects together.
- **Grade-11 baseline results should be withheld.** They contradict every cross-sectional measurement of the same cohort and pile up at the ceiling of the distribution.
- **Within-year growth is comparable across student groups**, with students with disabilities the substantial exception.
- **Catch-up / keep-up rates are not comparable across grades** without their target horizon, which runs from three years at grade 4 to none at grade 8.

::: {.notes}
Land the bottom line, and be explicit that two conclusions changed from the previous revision: mathematics is reframed from concern to recovery, and grade-11 baseline reporting moves from caveated to withheld. Both changes came from committee feedback, and both were checkable against data already in hand. Credit that.
:::


## Questions for the TAC {.smaller}

::: {.key-insight-box}
::: {.key-insight-box-title}
Where the committee's guidance is still needed
:::
1. **Baseline norm maintenance &mdash; genuinely open.** The written comments point both ways: that the 2017&ndash;19 window may not represent normal growth and should be reset, perhaps pooling several two-year periods; and that re-anchoring destroys the very instrument that makes drift and recovery detectable. Both are right about something. Is the resolution a **dual track** &mdash; a fixed monitoring series plus a re-established reporting norm &mdash; and if so, on what cadence?
2. **Is the ELA growth pattern the same phenomenon as the ELA equating pattern?** This deck establishes that ELA is down on both conditional and unconditional measures. Whether that connects to the one-directional pattern reported in the MCAS ELA equating is checkable, and needs the Center and Cognia looking at both sets of numbers together.
3. **Form effects, designed once.** Rhode Island's form-effects question and Cognia's validation question appear to be the same study. Should RIDE commission a single design, and approach Massachusetts jointly &mdash; population invariance being a requirement of equating rather than a courtesy?
4. **Reporting guardrails.** Proposed: never publish a baseline SGP without the cohort SGP beside it; do not use baseline SGPs in accountability while the drift question is open; withhold grade-11 baseline entirely. Are these the right three?
:::

::: {.notes}
Leave this up and open the floor. Question 1 is the one that most needs resolving, and the deck deliberately does not take a position because the committee's own written comments contain both positions; ask which was intended. Questions 2 and 3 are commissioning decisions rather than analytic ones. Question 4 is a check on guardrails this revision has already adopted.
:::


## Thank you {.center}

::: {style="text-align: center; padding-top: 1em;"}
::: {.conference-badge}
RIDE Technical Advisory Committee
:::

**Rhode Island 2026 SGP Analyses**

::: {style="font-size: 0.7em; color: #7c7c7c;"}
Damian W. Betebenner &middot; National Center for the Improvement of Educational Assessment
:::
:::

::: {.notes}
Close and thank the committee, and mean it: this revision exists because the feedback was specific enough to act on, and two of its conclusions changed as a result. Offer to circulate the deck and the pipeline, and note that further feedback can be turned around the same way &mdash; the deck rebuilds from the analysis rather than being rewritten.
:::
